---
title: "R Notebook"
output: html_notebook
---

```{r}
#install.packages("readxl")
#install.packages("skimr")
#install.packages("patchwork")
#install.packages("tseries")
#install.packages("forecast")
#install.packages("vars")
#install.packages("dynlm")
library(dynlm)
library(readxl)
library(data.table)
library(dplyr)
library(tidyr)
library(ggplot2)
library(skimr)
library(patchwork)
library(tseries)
library(forecast)
library(vars)
```
# Data Preprocessing 

```{r}



cropurl <- "https://catalog.ourworldindata.org/explorers/agriculture/latest/crop_yields/crop_yields.csv"
crop_yields <- fread(cropurl)
rainearly <- read_excel("Rain1950-2014.xlsx")
rainlate <- read_excel("Rain2015-2024.xlsx")
temp <- read_excel("TempCelsius.xlsx")

# View the first few rows
head(crop_yields)

rainearly
rainlate
temp <- temp %>% 
  rename(country = Name)
temp

```
## Precipitation Table formatting
```{r}

precipitation <- merge(rainearly, rainlate, by = "code", all = FALSE)
colnames(precipitation) <- sub("-07", "", colnames(precipitation))

precipitation <- precipitation %>%
  rename(country = name.x)

precipitation <- precipitation %>% 
  dplyr::select(-code)


precipitation[, as.character(1950:2024)] <- lapply(precipitation[, as.character(1950:2024)], as.integer) #Gotta include as.character cause R isn't smart and doesn't understand a number (grrrr)

precipitation
```





## Pivoting so Countries are the majority of columns and the years are just one year (Transpose was googled)
```{r}
filtered_prec <- setNames(data.frame(t(precipitation[,-1])), precipitation[,1])

filtered_prec <- cbind(year = as.integer(rownames(filtered_prec)), filtered_prec)
rownames(filtered_prec) <- NULL # Cool lil trick

filtered_prec[-1] <- lapply(filtered_prec[-1], as.integer)
filtered_prec


```
## Same process from Temp just copied pasted
```{r}
temp <- temp %>% 
  dplyr::select(-Code)


temp[, as.character(1901:2023)] <- lapply(temp[, as.character(1901:2023)], as.double)

temp
```

## Same Process as Precipitation with a few minor changes (Wanted to see hour extracting first worked)
```{r}
# This made it much easier, but wanted to keep my prior struggles up
country_names <- temp$country


filtered_temp <- as.data.frame(t(temp[, -1]))


colnames(filtered_temp) <- country_names


filtered_temp <- cbind(
  year = as.integer(rownames(filtered_temp)),
  filtered_temp)
rownames(filtered_temp) <- NULL

filtered_temp

```

## Creating the 5 main yield tables for the selected Countries
```{r}
# Select relevant columns
wheat_data <- crop_yields %>%
  dplyr::select(country, year, wheat_yield)

rice_data <- crop_yields %>%
  dplyr::select(country, year, rice_yield)

potato_data <- crop_yields %>%
  dplyr::select(country, year, potato_yield)

cotton_data <- crop_yields %>%
  dplyr::select(country, year, cotton_yield)

maize_data <- crop_yields %>%
  dplyr::select(country, year, maize_yield)


# It's Pivot Time (Insert Friends Meme)

wheat_data <- wheat_data %>%
  pivot_wider(names_from = country, values_from = wheat_yield) %>% 
  dplyr::select('year','Australia', 'Canada', 'Germany', 'France', 'India', 'Mexico', 'Poland', 'United States')

rice_data <- rice_data %>%
  pivot_wider(names_from = country, values_from = rice_yield) %>% 
  dplyr::select('year','Australia', 'Canada', 'Germany', 'France', 'India', 'Mexico', 'Poland', 'United States')

potato_data <- potato_data %>%
  pivot_wider(names_from = country, values_from = potato_yield) %>% 
  dplyr::select('year','Australia', 'Canada', 'Germany', 'France', 'India', 'Mexico', 'Poland', 'United States')

cotton_data <- cotton_data %>%
  pivot_wider(names_from = country, values_from = cotton_yield) %>% 
  dplyr::select('year','Australia', 'Canada', 'Germany', 'France', 'India', 'Mexico', 'Poland', 'United States')

maize_data <- maize_data %>%
  pivot_wider(names_from = country, values_from = maize_yield) %>% 
  dplyr::select('year','Australia', 'Canada', 'Germany', 'France', 'India', 'Mexico', 'Poland', 'United States')

wheat_data
rice_data
potato_data
cotton_data
maize_data
```



## Final Tables

```{r}

filtered_temp # Measured in Celcius
filtered_prec # Measured in MM
wheat_data #Yields are measured in tonnes per hectare (Whatever tf that means)
rice_data #Yields are measured in tonnes per hectare
potato_data #Yields are measured in tonnes per hectare
cotton_data #Yields are measured in tonnes per hectare
maize_data #Yields are measured in tonnes per hectare
```
## Simple EDA

```{r}


prec_long <- filtered_prec %>%
  pivot_longer(cols = -year, names_to = "country", values_to = "Precipitation") %>%
  drop_na() %>%
  filter(year >= 1960)

temp_long <- filtered_temp %>%
  pivot_longer(cols = -year, names_to = "country", values_to = "Temperature") %>%
  drop_na() %>%
  filter(year >= 1960)

wheat_long <- wheat_data %>%
  pivot_longer(cols = -year, names_to = "country", values_to = "Wheat_Yield") %>%
  drop_na() %>%
  filter(year >= 1960)

rice_long <- rice_data %>%
  pivot_longer(cols = -year, names_to = "country", values_to = "Rice_Yield") %>%
  drop_na() %>%
  filter(year >= 1960)

potato_long <- potato_data %>%
  pivot_longer(cols = -year, names_to = "country", values_to = "Potato_Yield") %>%
  drop_na() %>%
  filter(year >= 1960)

cotton_long <- cotton_data %>%
  pivot_longer(cols = -year, names_to = "country", values_to = "Cotton_Yield") %>%
  drop_na() %>%
  filter(year >= 1960)

maize_long <- maize_data %>%
  pivot_longer(cols = -year, names_to = "country", values_to = "Maize_Yield") %>%
  drop_na() %>%
  filter(year >= 1960)


```


```{r}
CountryPrecipitation <- ggplot(prec_long, aes(x = year, y = as.numeric(Precipitation), color = country)) +
  geom_line() +
  labs(title = "Annual Precipitation by Country (1961-2021)",
       x = "Year",
       y = "Precipitation (mm)",
       color = "Country") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(1961, 2021, by = 5)) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))
CountryPrecipitation
```

```{r}
CountryTemp <- ggplot(temp_long, aes(x = year, y = as.numeric(Temperature), color = country)) +
  geom_line() +
  labs(title = "Annual Temperature by Country (1900-2021)",
       x = "Year",
       y = "Temperature (Celcius)",
       color = "Country") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(1900, 2021, by = 5)) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))
CountryTemp
```
```{r}
annualWheat <- ggplot(wheat_long, aes(x = year, y = as.numeric(Wheat_Yield), color = country)) +
  geom_line() +
  labs(title = "Annual Wheat Yield  (1961-2021)",
       x = "Year",
       y = "Wheat_Yield (Tonnes per Hectare)",
       color = "Country") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(1961, 2021, by = 3)) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))
annualWheat
```
```{r}
annualRice <- ggplot(rice_long, aes(x = year, y = as.numeric(Rice_Yield), color = country)) +
  geom_line() +
  labs(title = "Annual Rice Yield  (1961-2021)",
       x = "Year",
       y = "Rice_Yield (Tonnes per Hectare)",
       color = "Country") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(1961, 2021, by = 3)) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))
annualRice
```

```{r}
annualPotato <- ggplot(potato_long , aes(x = year, y = as.numeric(Potato_Yield), color = country)) +
  geom_line() +
  labs(title = "Annual Potato Yield  (1961-2021)",
       x = "Year",
       y = "Potato_Yield (Tonnes per Hectare)",
       color = "Country") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(1961, 2021, by = 3)) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))
annualPotato
```

```{r}
annualCotton <- ggplot(cotton_long , aes(x = year, y = as.numeric(Cotton_Yield), color = country)) +
  geom_line() +
  labs(title = "Annual Cotton Yield  (1961-2021)",
       x = "Year",
       y = "Cotton_Yield (Tonnes per Hectare)",
       color = "Country") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(1961, 2021, by = 3)) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))
annualCotton
```

```{r}
annualMaize <- ggplot(maize_long , aes(x = year, y = as.numeric(Maize_Yield), color = country)) +
  geom_line() +
  labs(title = "Annual Maize Yield  (1961-2021)",
       x = "Year",
       y = "Maize_Yield (Tonnes per Hectare)",
       color = "Country") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(1961, 2021, by = 3)) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))
annualMaize
```




```{r}

CountryPrecipitation + plot_spacer() + CountryTemp

```

```{r}

library(tseries)

adf_temp <- adf.test(temp_long$Temperature, alternative = "stationary")
adf_prec <- adf.test(prec_long$Precipitation, alternative = "stationary")
adf_wheat <- adf.test(wheat_long$Wheat_Yield, alternative = "stationary")
adf_rice <- adf.test(rice_long$Rice_Yield, alternative = "stationary")
adf_potato <- adf.test(potato_long$Potato_Yield, alternative = "stationary")
adf_cotton <- adf.test(cotton_long$Cotton_Yield, alternative = "stationary")
adf_maize <- adf.test(maize_long$Maize_Yield, alternative = "stationary")

print(adf_temp)
print(adf_prec)
print(adf_wheat)
print(adf_rice)
print(adf_potato) # Non-Stationary so it might have Trend or Seasonality
print(adf_cotton)
print(adf_maize)

```

## Differencing Non-Stationary Models

```{r}

potato_long$Potato_Yield_diff <- c(NA, diff(potato_long$Potato_Yield))

potato_long_diff <- potato_long$Potato_Yield_diff[!is.na(potato_long$Potato_Yield_diff)]

# Perform the ADF test on the differenced data
adf.test(potato_long_diff, alternative = "stationary")
```
```{r}

merged_data <- merge(wheat_long, temp_long, by = c("country", "year"))
merged_data <- merge(merged_data, prec_long, by = c("country", "year"))
merged_data <- merge(merged_data, rice_long, by = c("country", "year"))
merged_data <- merge(merged_data, potato_long, by = c("country", "year"))
merged_data <- merge(merged_data, cotton_long, by = c("country", "year"))
merged_data <- merge(merged_data, maize_long, by = c("country", "year"))
head(merged_data)

```


```{r}


# ACF and PACF for Temperature
acf(temp_long$Temperature, main="ACF of Temperature")
pacf(temp_long$Temperature, main="PACF of Temperature")

# ACF and PACF for Precipitation
acf(prec_long$Precipitation, main="ACF of Precipitation")
pacf(prec_long$Precipitation, main="PACF of Precipitation")

# ACF and PACF for Wheat Yield
acf(wheat_long$Wheat_Yield, main="ACF of Wheat Yield")
pacf(wheat_long$Wheat_Yield, main="PACF of Wheat Yield")

# ACF and PACF for Rice Yield
acf(rice_long$Rice_Yield, main="ACF of Rice Yield")
pacf(rice_long$Rice_Yield, main="PACF of Rice Yield")

# ACF and PACF for Potato Yield
acf(potato_long$Potato_Yield, main="ACF of Potato Yield")
pacf(potato_long$Potato_Yield, main="PACF of Potato Yield")

# ACF and PACF for Cotton Yield
acf(cotton_long$Cotton_Yield, main="ACF of Cotton Yield")
pacf(cotton_long$Cotton_Yield, main="PACF of Cotton Yield")

# ACF and PACF for Maize Yield
acf(maize_long$Maize_Yield, main="ACF of Maize Yield")
pacf(maize_long$Maize_Yield, main="PACF of Maize Yield")


```

```{r}

ccf(temp_long$Temperature, wheat_long$Wheat_Yield, main="Cross-Correlation: Temperature vs Wheat Yield")
ccf(prec_long$Precipitation, wheat_long$Wheat_Yield, main="Cross-Correlation: Precipitation vs Wheat Yield")

ccf(temp_long$Temperature, rice_long$Rice_Yield, main="Cross-Correlation: Temperature vs Rice Yield")
ccf(prec_long$Precipitation, rice_long$Rice_Yield, main="Cross-Correlation: Precipitation vs Rice Yield")

ccf(temp_long$Temperature, potato_long$Potato_Yield, main="Cross-Correlation: Temperature vs Potato Yield")
ccf(prec_long$Precipitation, potato_long$Potato_Yield, main="Cross-Correlation: Precipitation vs Potato Yield")

ccf(temp_long$Temperature, cotton_long$Cotton_Yield, main="Cross-Correlation: Temperature vs Cotton Yield")
ccf(prec_long$Precipitation, cotton_long$Cotton_Yield, main="Cross-Correlation: Precipitation vs Cotton Yield")

ccf(temp_long$Temperature, maize_long$Maize_Yield, main="Cross-Correlation: Temperature vs Maize Yield")
ccf(prec_long$Precipitation, maize_long$Maize_Yield, main="Cross-Correlation: Precipitation vs Maize Yield")



```
```{r}

# Create lagged variables for Temperature
merged_data$Temp_Lag1 <- lag(merged_data$Temperature, 1)
merged_data$Temp_Lag2 <- lag(merged_data$Temperature, 2)
merged_data$Temp_Lag3 <- lag(merged_data$Temperature, 3)

# Check the first few rows of merged_data
merged_data
```


```{r}
# Create lagged variables for Temperature (lag 1 to 3 years)
merged_data$Temp_Lag1 <- lag(merged_data$Temperature, 1)
merged_data$Temp_Lag2 <- lag(merged_data$Temperature, 2)
merged_data$Temp_Lag3 <- lag(merged_data$Temperature, 3)

# Create lagged variables for Precipitation (lag 1 to 3 years)
merged_data$Prec_Lag1 <- lag(merged_data$Precipitation, 1)
merged_data$Prec_Lag2 <- lag(merged_data$Precipitation, 2)
merged_data$Prec_Lag3 <- lag(merged_data$Precipitation, 3)

# Create lagged variables for Wheat Yield (lag 1 to 3 years)
merged_data$Wheat_Yield_Lag1 <- lag(merged_data$Wheat_Yield, 1)
merged_data$Wheat_Yield_Lag2 <- lag(merged_data$Wheat_Yield, 2)
merged_data$Wheat_Yield_Lag3 <- lag(merged_data$Wheat_Yield, 3)

# Create lagged variables for Rice Yield (lag 1 to 3 years)
merged_data$Rice_Yield_Lag1 <- lag(merged_data$Rice_Yield, 1)
merged_data$Rice_Yield_Lag2 <- lag(merged_data$Rice_Yield, 2)
merged_data$Rice_Yield_Lag3 <- lag(merged_data$Rice_Yield, 3)

# Create lagged variables for Potato Yield (lag 1 to 3 years)
merged_data$Potato_Yield_Lag1 <- lag(merged_data$Potato_Yield, 1)
merged_data$Potato_Yield_Lag2 <- lag(merged_data$Potato_Yield, 2)
merged_data$Potato_Yield_Lag3 <- lag(merged_data$Potato_Yield, 3)

# Create lagged variables for Cotton Yield (lag 1 to 3 years)
merged_data$Cotton_Yield_Lag1 <- lag(merged_data$Cotton_Yield, 1)
merged_data$Cotton_Yield_Lag2 <- lag(merged_data$Cotton_Yield, 2)
merged_data$Cotton_Yield_Lag3 <- lag(merged_data$Cotton_Yield, 3)

# Create lagged variables for Maize Yield (lag 1 to 3 years)
merged_data$Maize_Yield_Lag1 <- lag(merged_data$Maize_Yield, 1)
merged_data$Maize_Yield_Lag2 <- lag(merged_data$Maize_Yield, 2)
merged_data$Maize_Yield_Lag3 <- lag(merged_data$Maize_Yield, 3)

# Check the first few rows of the merged data with lags
head(merged_data)


```


```{r}
library(dynlm)

# Wheat model (already provided)
dlm_wheat <- dynlm(
  Wheat_Yield ~ Temperature + Temp_Lag1 + Temp_Lag2 + Temp_Lag3 + 
  Precipitation + Prec_Lag1 + Prec_Lag2 + Prec_Lag3, 
  data = merged_data
)

# Rice model
dlm_rice <- dynlm(
  Rice_Yield ~ Temperature + Temp_Lag1 + Temp_Lag2 + Temp_Lag3 + 
  Precipitation + Prec_Lag1 + Prec_Lag2 + Prec_Lag3, 
  data = merged_data
)

# Potato model
dlm_potato <- dynlm(
  Potato_Yield ~ Temperature + Temp_Lag1 + Temp_Lag2 + Temp_Lag3 + 
  Precipitation + Prec_Lag1 + Prec_Lag2 + Prec_Lag3, 
  data = merged_data
)

# Cotton model
dlm_cotton <- dynlm(
  Cotton_Yield ~ Temperature + Temp_Lag1 + Temp_Lag2 + Temp_Lag3 + 
  Precipitation + Prec_Lag1 + Prec_Lag2 + Prec_Lag3, 
  data = merged_data
)

# Maize model
dlm_maize <- dynlm(
  Maize_Yield ~ Temperature + Temp_Lag1 + Temp_Lag2 + Temp_Lag3 + 
  Precipitation + Prec_Lag1 + Prec_Lag2 + Prec_Lag3, 
  data = merged_data
)

crop_models <- list(
  Wheat = dlm_wheat,
  Rice = dlm_rice,
  Potato = dlm_potato,
  Cotton = dlm_cotton,
  Maize = dlm_maize
)

lapply(crop_models, summary)


```


# Let's Try some Prediction
```{r}

set.seed(123)
filtered_data <- merged_data %>% filter(year >= 1990) %>% arrange(year)
n_rows <- nrow(filtered_data)
train_rows <- round(0.80 * n_rows)
train_data <- filtered_data %>% slice(1:train_rows)
test_data <- filtered_data %>% slice((train_rows + 1):n_rows)

# Fit a linear regression model on the training data
model <- lm(Wheat_Yield ~ Temperature + Temp_Lag1 + Temp_Lag2 + Temp_Lag3 +
                                  Precipitation + Prec_Lag1 + Prec_Lag2 + Prec_Lag3, 
            data = train_data)

# Summary of the model
summary(model)




```

```{r}
# Predict on test data
predictions <- predict(model, newdata = test_data)

# Add the predictions to the test data for comparison
test_data$predictions <- predictions

```


```{r}
# Plot actual vs predicted values
ggplot(test_data) +
  geom_line(aes(x = year, y = Wheat_Yield, color = "Actual"), linewidth = 1) +
  geom_line(aes(x = year, y = predictions, color = "Predicted"), linewidth = 1) +
  scale_color_manual(values = c("Actual" = "blue", "Predicted" = "red")) +
  labs(title = "Actual vs Predicted Wheat Yield (Test Set)", x = "Year", y = "Wheat Yield (tons/ha)") +
  theme_minimal() +
  theme(legend.position = "bottom")

```

```{r}
# Calculate RMSE (Root Mean Squared Error)
rmse <- sqrt(mean((test_data$Wheat_Yield - test_data$predictions)^2))
print(paste("RMSE: ", round(rmse, 2)))

# Calculate MAE (Mean Absolute Error)
mae <- mean(abs(test_data$Wheat_Yield - test_data$predictions))
print(paste("MAE: ", round(mae, 2)))

```








