---
title: "Analysis of Temperature and Precipitation Effects on Agricultural Yields"
output:
  html_notebook:
    toc: yes
    toc_float: yes
    theme: united
  html_document:
    toc: yes
    df_print: paged
date: "2025-03-31"
---

# Introduction

This report investigates the critical research question: How do temperature and precipitation changes affect agricultural yields? Understanding these relationships is vital for several reasons:

* Food Security: With global population projected to only increase, optimizing agricultural productivity is essential.

* Climate Change Adaptation: Farmers and policymakers need data-driven insights to adapt to changing climate patterns.

* Economic Stability: Agriculture contributes significantly to many nations' GDPs and employment.

The analysis focuses on five key crops (wheat, rice, potatoes, cotton, and maize) across eight countries (Australia, Canada, Germany, France, India, Mexico, Poland, and the United States) from 1961-2021.

* This is a small selection from 20+ crops and all countries in the world


# PreProcessing


## Import Packages and Libraries
```{r}
#install.packages("readxl")
#install.packages("skimr")
#install.packages("patchwork")
#install.packages("tseries")
#install.packages("forecast")
#install.packages("vars")
#install.packages("dynlm")
library(dynlm)
library(readxl)
library(data.table)
library(dplyr)
library(tidyr)
library(ggplot2)
library(skimr)
library(patchwork)
library(tseries)
library(forecast)
library(vars)
library(tseries)
```


## Data Preprocessing

```{r}



cropurl <- "https://catalog.ourworldindata.org/explorers/agriculture/latest/crop_yields/crop_yields.csv"
crop_yields <- fread(cropurl)
rainearly <- read_excel("Rain1950-2014.xlsx")
rainlate <- read_excel("Rain2015-2024.xlsx")
temp <- read_excel("TempCelsius.xlsx")


crop_yields
rainearly
rainlate
temp <- temp %>% 
  rename(country = Name)
temp

```


## Precipitation Table formatting
```{r}

precipitation <- merge(rainearly, rainlate, by = "code", all = FALSE)
colnames(precipitation) <- sub("-07", "", colnames(precipitation))

precipitation <- precipitation %>%
  rename(country = name.x)

precipitation <- precipitation %>% 
  dplyr::select(-code)


precipitation[, as.character(1950:2024)] <- lapply(precipitation[, as.character(1950:2024)], as.integer) #Gotta include as.character cause R isn't smart and doesn't understand a number (grrrr)

precipitation
```


## Pivoting the Countries via Transpose
```{r}
filtered_prec <- setNames(data.frame(t(precipitation[,-1])), precipitation[,1])

filtered_prec <- cbind(year = as.integer(rownames(filtered_prec)), filtered_prec)
rownames(filtered_prec) <- NULL # Cool lil trick

filtered_prec[-1] <- lapply(filtered_prec[-1], as.integer)
filtered_prec

```


## Temperature Table Formatting
```{r}
temp <- temp %>% 
  dplyr::select(-Code)


temp[, as.character(1901:2023)] <- lapply(temp[, as.character(1901:2023)], as.double)

temp
```

## Temperature Table Formatting pt.2
```{r}
# This made it much easier, but wanted to keep my prior struggles up
country_names <- temp$country


filtered_temp <- as.data.frame(t(temp[, -1]))


colnames(filtered_temp) <- country_names


filtered_temp <- cbind(
  year = as.integer(rownames(filtered_temp)),
  filtered_temp)
rownames(filtered_temp) <- NULL

filtered_temp

```

## Yield Tables and Ross is Pivoting
```{r}

# Quick note to self, use dplyr because it seems select() has some issues with another package

wheat_data <- crop_yields %>%
  dplyr::select(country, year, wheat_yield)

rice_data <- crop_yields %>%
  dplyr::select(country, year, rice_yield)

potato_data <- crop_yields %>%
  dplyr::select(country, year, potato_yield)

cotton_data <- crop_yields %>%
  dplyr::select(country, year, cotton_yield)

maize_data <- crop_yields %>%
  dplyr::select(country, year, maize_yield)


# It's Pivot Time (Insert Friends Meme)

wheat_data <- wheat_data %>%
  pivot_wider(names_from = country, values_from = wheat_yield) %>% 
  dplyr::select('year','Australia', 'Canada', 'Germany', 'France', 'India', 'Mexico', 'Poland', 'United States')

rice_data <- rice_data %>%
  pivot_wider(names_from = country, values_from = rice_yield) %>% 
  dplyr::select('year','Australia', 'Canada', 'Germany', 'France', 'India', 'Mexico', 'Poland', 'United States')

potato_data <- potato_data %>%
  pivot_wider(names_from = country, values_from = potato_yield) %>% 
  dplyr::select('year','Australia', 'Canada', 'Germany', 'France', 'India', 'Mexico', 'Poland', 'United States')

cotton_data <- cotton_data %>%
  pivot_wider(names_from = country, values_from = cotton_yield) %>% 
  dplyr::select('year','Australia', 'Canada', 'Germany', 'France', 'India', 'Mexico', 'Poland', 'United States')

maize_data <- maize_data %>%
  pivot_wider(names_from = country, values_from = maize_yield) %>% 
  dplyr::select('year','Australia', 'Canada', 'Germany', 'France', 'India', 'Mexico', 'Poland', 'United States')

wheat_data
rice_data
potato_data
cotton_data
maize_data
```

## Final Tables

```{r}

filtered_temp # Measured in Celcius
filtered_prec # Measured in MM
wheat_data #Yields are measured in tonnes per hectare (Whatever tf that means)
rice_data #Yields are measured in tonnes per hectare
potato_data #Yields are measured in tonnes per hectare
cotton_data #Yields are measured in tonnes per hectare
maize_data #Yields are measured in tonnes per hectare
```


## Pivoting (Again)

```{r}

# After a few hours of testing I saw I NEEDED TO PIVOT AGAIN

prec_long <- filtered_prec %>%
  pivot_longer(cols = -year, names_to = "country", values_to = "Precipitation") %>%
  drop_na() %>%
  filter(year >= 1960)

temp_long <- filtered_temp %>%
  pivot_longer(cols = -year, names_to = "country", values_to = "Temperature") %>%
  drop_na() %>%
  filter(year >= 1960)

wheat_long <- wheat_data %>%
  pivot_longer(cols = -year, names_to = "country", values_to = "Wheat_Yield") %>%
  drop_na() %>%
  filter(year >= 1960)

rice_long <- rice_data %>%
  pivot_longer(cols = -year, names_to = "country", values_to = "Rice_Yield") %>%
  drop_na() %>%
  filter(year >= 1960)

potato_long <- potato_data %>%
  pivot_longer(cols = -year, names_to = "country", values_to = "Potato_Yield") %>%
  drop_na() %>%
  filter(year >= 1960)

cotton_long <- cotton_data %>%
  pivot_longer(cols = -year, names_to = "country", values_to = "Cotton_Yield") %>%
  drop_na() %>%
  filter(year >= 1960)

maize_long <- maize_data %>%
  pivot_longer(cols = -year, names_to = "country", values_to = "Maize_Yield") %>%
  drop_na() %>%
  filter(year >= 1960)


```


# Testing and Insights

The methodology employed several time series techniques:

* Stationarity Testing: Augmented Dickey-Fuller tests to check for non-stationarity

* Autocorrelation Analysis: ACF/PACF plots to identify temporal patterns

* Cross-Correlation: CCF plots to examine lead-lag relationships 

* Dynamic Linear Models (DLM): Regression models incorporating lagged effects

## EDA

```{r}

summary(prec_long)
summary(temp_long)
summary(wheat_long)
summary(rice_long)
summary(potato_long)
summary(cotton_long)
summary(maize_long)
```

### Annual Precipitation by Country

```{r}
CountryPrecipitation <- ggplot(prec_long, aes(x = year, y = as.numeric(Precipitation), color = country)) +
  geom_line() +
  labs(title = "Annual Precipitation by Country (1961-2021)",
       x = "Year",
       y = "Precipitation (mm)",
       color = "Country") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(1961, 2021, by = 5)) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))
CountryPrecipitation
```


### Annual Temperature by Country

```{r}
CountryTemp <- ggplot(temp_long, aes(x = year, y = as.numeric(Temperature), color = country)) +
  geom_line() +
  labs(title = "Annual Temperature by Country (1900-2021)",
       x = "Year",
       y = "Temperature (Celcius)",
       color = "Country") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(1900, 2021, by = 5)) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))
CountryTemp
```


### Annual Wheat Yield
```{r}
annualWheat <- ggplot(wheat_long, aes(x = year, y = as.numeric(Wheat_Yield), color = country)) +
  geom_line() +
  labs(title = "Annual Wheat Yield  (1961-2021)",
       x = "Year",
       y = "Wheat_Yield (Tonnes per Hectare)",
       color = "Country") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(1961, 2021, by = 3)) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))
annualWheat
```
```{r}
annualRice <- ggplot(rice_long, aes(x = year, y = as.numeric(Rice_Yield), color = country)) +
  geom_line() +
  labs(title = "Annual Rice Yield  (1961-2021)",
       x = "Year",
       y = "Rice_Yield (Tonnes per Hectare)",
       color = "Country") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(1961, 2021, by = 3)) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))
annualRice
```

### Annual Potato Yield

```{r}
annualPotato <- ggplot(potato_long , aes(x = year, y = as.numeric(Potato_Yield), color = country)) +
  geom_line() +
  labs(title = "Annual Potato Yield  (1961-2021)",
       x = "Year",
       y = "Potato_Yield (Tonnes per Hectare)",
       color = "Country") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(1961, 2021, by = 3)) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))
annualPotato
```

### Annual Cotton Yield

```{r}
annualCotton <- ggplot(cotton_long , aes(x = year, y = as.numeric(Cotton_Yield), color = country)) +
  geom_line() +
  labs(title = "Annual Cotton Yield  (1961-2021)",
       x = "Year",
       y = "Cotton_Yield (Tonnes per Hectare)",
       color = "Country") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(1961, 2021, by = 3)) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))
annualCotton
```


## Annual Maize Yield


```{r}
annualMaize <- ggplot(maize_long , aes(x = year, y = as.numeric(Maize_Yield), color = country)) +
  geom_line() +
  labs(title = "Annual Maize Yield  (1961-2021)",
       x = "Year",
       y = "Maize_Yield (Tonnes per Hectare)",
       color = "Country") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(1961, 2021, by = 3)) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))
annualMaize
```

## ADF test

```{r}
adf_temp <- adf.test(temp_long$Temperature, alternative = "stationary")
adf_prec <- adf.test(prec_long$Precipitation, alternative = "stationary")
adf_wheat <- adf.test(wheat_long$Wheat_Yield, alternative = "stationary")
adf_rice <- adf.test(rice_long$Rice_Yield, alternative = "stationary")
adf_potato <- adf.test(potato_long$Potato_Yield, alternative = "stationary")
adf_cotton <- adf.test(cotton_long$Cotton_Yield, alternative = "stationary")
adf_maize <- adf.test(maize_long$Maize_Yield, alternative = "stationary")

print(adf_temp)
print(adf_prec)
print(adf_wheat)
print(adf_rice)
print(adf_potato) # Non-Stationary so it might have Trend or Seasonality
print(adf_cotton)
print(adf_maize)

```

### Differencing Non-Stationary Models

```{r}

potato_long$Potato_Yield_diff <- c(NA, diff(potato_long$Potato_Yield))

potato_long_diff <- potato_long$Potato_Yield_diff[!is.na(potato_long$Potato_Yield_diff)]

# Potatos man, why aren't you stationary
adf.test(potato_long_diff, alternative = "stationary")
```
```{r}

merged_data <- merge(wheat_long, temp_long, by = c("country", "year"))
merged_data <- merge(merged_data, prec_long, by = c("country", "year"))
merged_data <- merge(merged_data, rice_long, by = c("country", "year"))
merged_data <- merge(merged_data, potato_long, by = c("country", "year"))
merged_data <- merge(merged_data, cotton_long, by = c("country", "year"))
merged_data <- merge(merged_data, maize_long, by = c("country", "year"))
head(merged_data)

```

## ACF and PACF

```{r}

# Thank goodness for the guy who invinted copy/paste

acf(temp_long$Temperature, main="ACF of Temperature")
pacf(temp_long$Temperature, main="PACF of Temperature")


acf(prec_long$Precipitation, main="ACF of Precipitation")
pacf(prec_long$Precipitation, main="PACF of Precipitation")


acf(wheat_long$Wheat_Yield, main="ACF of Wheat Yield")
pacf(wheat_long$Wheat_Yield, main="PACF of Wheat Yield")


acf(rice_long$Rice_Yield, main="ACF of Rice Yield")
pacf(rice_long$Rice_Yield, main="PACF of Rice Yield")


acf(potato_long$Potato_Yield, main="ACF of Potato Yield")
pacf(potato_long$Potato_Yield, main="PACF of Potato Yield")


acf(cotton_long$Cotton_Yield, main="ACF of Cotton Yield")
pacf(cotton_long$Cotton_Yield, main="PACF of Cotton Yield")


acf(maize_long$Maize_Yield, main="ACF of Maize Yield")
pacf(maize_long$Maize_Yield, main="PACF of Maize Yield")


```

## Cross-Correlation Plots

```{r}

ccf(temp_long$Temperature, wheat_long$Wheat_Yield, main="Cross-Correlation: Temperature vs Wheat Yield")
ccf(prec_long$Precipitation, wheat_long$Wheat_Yield, main="Cross-Correlation: Precipitation vs Wheat Yield")

ccf(temp_long$Temperature, rice_long$Rice_Yield, main="Cross-Correlation: Temperature vs Rice Yield")
ccf(prec_long$Precipitation, rice_long$Rice_Yield, main="Cross-Correlation: Precipitation vs Rice Yield")

ccf(temp_long$Temperature, potato_long$Potato_Yield, main="Cross-Correlation: Temperature vs Potato Yield")
ccf(prec_long$Precipitation, potato_long$Potato_Yield, main="Cross-Correlation: Precipitation vs Potato Yield")

ccf(temp_long$Temperature, cotton_long$Cotton_Yield, main="Cross-Correlation: Temperature vs Cotton Yield")
ccf(prec_long$Precipitation, cotton_long$Cotton_Yield, main="Cross-Correlation: Precipitation vs Cotton Yield")

ccf(temp_long$Temperature, maize_long$Maize_Yield, main="Cross-Correlation: Temperature vs Maize Yield")
ccf(prec_long$Precipitation, maize_long$Maize_Yield, main="Cross-Correlation: Precipitation vs Maize Yield")



```

## Creating the Lags Yo

```{r}

# 3 years of lagged variables cause I thought that was about right

merged_data$Temp_Lag1 <- lag(merged_data$Temperature, 1)
merged_data$Temp_Lag2 <- lag(merged_data$Temperature, 2)
merged_data$Temp_Lag3 <- lag(merged_data$Temperature, 3)


merged_data$Prec_Lag1 <- lag(merged_data$Precipitation, 1)
merged_data$Prec_Lag2 <- lag(merged_data$Precipitation, 2)
merged_data$Prec_Lag3 <- lag(merged_data$Precipitation, 3)


merged_data$Wheat_Yield_Lag1 <- lag(merged_data$Wheat_Yield, 1)
merged_data$Wheat_Yield_Lag2 <- lag(merged_data$Wheat_Yield, 2)
merged_data$Wheat_Yield_Lag3 <- lag(merged_data$Wheat_Yield, 3)


merged_data$Rice_Yield_Lag1 <- lag(merged_data$Rice_Yield, 1)
merged_data$Rice_Yield_Lag2 <- lag(merged_data$Rice_Yield, 2)
merged_data$Rice_Yield_Lag3 <- lag(merged_data$Rice_Yield, 3)


merged_data$Potato_Yield_Lag1 <- lag(merged_data$Potato_Yield, 1)
merged_data$Potato_Yield_Lag2 <- lag(merged_data$Potato_Yield, 2)
merged_data$Potato_Yield_Lag3 <- lag(merged_data$Potato_Yield, 3)


merged_data$Cotton_Yield_Lag1 <- lag(merged_data$Cotton_Yield, 1)
merged_data$Cotton_Yield_Lag2 <- lag(merged_data$Cotton_Yield, 2)
merged_data$Cotton_Yield_Lag3 <- lag(merged_data$Cotton_Yield, 3)


merged_data$Maize_Yield_Lag1 <- lag(merged_data$Maize_Yield, 1)
merged_data$Maize_Yield_Lag2 <- lag(merged_data$Maize_Yield, 2)
merged_data$Maize_Yield_Lag3 <- lag(merged_data$Maize_Yield, 3)


head(merged_data)


```

## Dynamic Linear Models

```{r}


dlm_wheat <- dynlm(
  Wheat_Yield ~ Temperature + Temp_Lag1 + Temp_Lag2 + Temp_Lag3 + 
  Precipitation + Prec_Lag1 + Prec_Lag2 + Prec_Lag3, 
  data = merged_data
)


dlm_rice <- dynlm(
  Rice_Yield ~ Temperature + Temp_Lag1 + Temp_Lag2 + Temp_Lag3 + 
  Precipitation + Prec_Lag1 + Prec_Lag2 + Prec_Lag3, 
  data = merged_data
)


dlm_potato <- dynlm(
  Potato_Yield ~ Temperature + Temp_Lag1 + Temp_Lag2 + Temp_Lag3 + 
  Precipitation + Prec_Lag1 + Prec_Lag2 + Prec_Lag3, 
  data = merged_data
)


dlm_cotton <- dynlm(
  Cotton_Yield ~ Temperature + Temp_Lag1 + Temp_Lag2 + Temp_Lag3 + 
  Precipitation + Prec_Lag1 + Prec_Lag2 + Prec_Lag3, 
  data = merged_data
)


dlm_maize <- dynlm(
  Maize_Yield ~ Temperature + Temp_Lag1 + Temp_Lag2 + Temp_Lag3 + 
  Precipitation + Prec_Lag1 + Prec_Lag2 + Prec_Lag3, 
  data = merged_data
)

crop_models <- list(
  Wheat = dlm_wheat,
  Rice = dlm_rice,
  Potato = dlm_potato,
  Cotton = dlm_cotton,
  Maize = dlm_maize
)

lapply(crop_models, summary) # New tool I found, perty cool


```

# Results

## EDA

Crop Yields:

* Wheat: Steady increase (2-4x higher yields since 1960)

* Rice: Similar growth pattern, especially pronounced in India

* Potatoes: More volatile with recent stagnation in some countries

* Cotton: Significant increases in the US and India

* Maize: Consistent growth across most countries

## Lagged Effects (Cross Correlation)

The cross-correlation analysis showed:

* Temperature shows strongest present time and 1-year lag effects

* Precipitation effects often appear with 1-2 year lags

* Wheat and maize show the clearest climate relationships



## DLM findings

The findings were all very small and minute, but still aplicable:

* Current-year temperature has negative effects across all crops (Very Small)

* 1-year lagged temperature shows compensatory positive effects (Still Very Small)

* Precipitation effects are positive (Small Once Again)

* Wheat and maize models show strongest explanatory power (R^2 which is my favorite)


## Key Findings

Temperature Effects: 

* Current-year warming reduces yields, but following years often show compensatory growth, this suggest some crop adaptation.


Precipitation Timing: 

* The lagged precipitation effects indicate soil moisture retention impacts future growing seasons.


Crop Vulnerability: 

* Potatoes show the strongest negative temperature sensitivity, while wheat and maize demonstrate more climate resilience.



# Conclusion


## Policy Implications

1. Adaptive Farming: Encourage crop rotation strategies that account for 1-2 year climate lag effects

2. Irrigation Investment: Prioritize water management in regions showing precipitation declines

3. Breeding Programs: Focus on developing temperature-resilient varieties, especially for potatoes (since they aren't very temperature resilient)


## Limitations and Future Research

### Limitations: 

* TECHNOLOGY: There aren't really accounting for technological improvements

* Major Weather: Excludes extreme weather events


### Future directions:

* Incorporate monthly climate data

* Add soil quality and irrigation variables

* Analyze climate change scenario projections


This analysis demonstrates that climate variables indeed have significant lagged effects on agricultural productivity, with important variations across crop types. These findings can inform both short-term agricultural planning and long-term climate adaptation strategies. While not completely accurate, this analysis helps the reader and farmers understand the difference that planetary occurrences has on agricultural life. 


# For Fun Prediction (I tried)

## Let's Try some Prediction
```{r}

set.seed(123)
filtered_data <- merged_data %>% filter(year >= 1990) %>% arrange(year)
n_rows <- nrow(filtered_data)
train_rows <- round(0.80 * n_rows)
train_data <- filtered_data %>% slice(1:train_rows)
test_data <- filtered_data %>% slice((train_rows + 1):n_rows)


model <- lm(Wheat_Yield ~ Temperature + Temp_Lag1 + Temp_Lag2 + Temp_Lag3 +
                                  Precipitation + Prec_Lag1 + Prec_Lag2 + Prec_Lag3, 
            data = train_data)


summary(model)




```

## Predict on test data

```{r}

predictions <- predict(model, newdata = test_data)


test_data$predictions <- predictions

```

## Plot actual vs predicted values

```{r}

ggplot(test_data) +
  geom_line(aes(x = year, y = Wheat_Yield, color = "Actual"), linewidth = 1) +
  geom_line(aes(x = year, y = predictions, color = "Predicted"), linewidth = 1) +
  scale_color_manual(values = c("Actual" = "blue", "Predicted" = "red")) +
  labs(title = "Actual vs Predicted Wheat Yield (Test Set)", x = "Year", y = "Wheat Yield (tons/ha)") +
  theme_minimal() +
  theme(legend.position = "bottom")

```

## Calculate RMSE and MAE (Only One)

```{r}

rmse <- sqrt(mean((test_data$Wheat_Yield - test_data$predictions)^2))
print(paste("RMSE: ", round(rmse, 2)))


mae <- mean(abs(test_data$Wheat_Yield - test_data$predictions))
print(paste("MAE: ", round(mae, 2)))

```








